# Getting and Cleaning Data 

# Course Assignment

## Instructions

The purpose of this project is to demonstrate your ability to collect, work with, and clean a data set.

### *Review criteria*

* The submitted data set is tidy. 

* The Github repo contains the required scripts.

* GitHub contains a code book that modifies and updates the available codebooks with the data to indicate all the variables and summaries calculated, along with units, and any other relevant information.

* The README that explains the analysis files is clear and understandable.

* The work submitted for this project is the work of the student who submitted it.
 
### *Getting and Cleaning Data Course Project*

The purpose of this project is to demonstrate your ability to collect, work with, and clean a data set. 

The goal is to prepare tidy data that can be used for later analysis. 

You will be graded by your peers on a series of yes/no questions related to the project. 

You will be required to submit: 

* a tidy data set as described below, 

* a link to a Github repository with your script for performing the analysis, and 

* a code book that describes the variables, the data, and any transformations or work that you performed to clean up the data called CodeBook.md. 

You should also include a README.md in the repo with your scripts. This repo explains how all of the scripts work and how they are connected.

One of the most exciting areas in all of data science right now is wearable computing - see for example this article . 
Companies like Fitbit, Nike, and Jawbone Up are racing to develop the most advanced algorithms to attract new users. 
The data linked to from the course website represent data collected from the accelerometers from the Samsung Galaxy S smartphone. 

A full description is available at the site where the data was obtained: 
(http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones)
 
Here are the data for the project: 
(https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip)

You should create one R script called run_analysis.R that does the following. 
* Merges the training and the test sets to create one data set.

* Extracts only the measurements on the mean and standard deviation for each measurement. 

* Uses descriptive activity names to name the activities in the data set

* Appropriately labels the data set with descriptive variable names. 

* From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject.

*Good luck!*

**____________________________________________________________________________**

* Require "plyr", "dplyr" & "reshape2" packages

```{r message=FALSE, warning=FALSE}
library(plyr)
library(dplyr)
library(reshape2)
```

* Create variables to identify the data url and data zip file

```{r}
fileURL <- "https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip"
UCIHARDataset <- "./data/UCI HAR Dataset.zip"
```

* Check if 'data' directory exists
* Create 'data' if it does not

```{r}
if (!file.exists("data")) {
  dir.create("data")
}
```

* Check if 'UCIHARDataset' file exists
* Download file as 'UCIHARDataset' if it does not
* Extract files from the zipped folder
* Place files in 'data' directory

```{r}
if (!file.exists(UCIHARDataset)){
  download.file(fileURL, destfile = UCIHARDataset, mode = "wb")
  unzip(UCIHARDataset, exdir = "./data")
  dateDownloaded <- date()
}
```

* Read in features.txt, comprised of the 561 feature descriptions
* Rename the columns in the 'features' dataframe
* Extract the feature descriptions into 'feature_names' vector
* Tidy up the feature descriptions in 'feature_names' so they are easier to read and understand

```{r}
features <- read.table("./data/UCI HAR Dataset/features.txt", stringsAsFactors = F)
features <- rename(features, feature_no = V1, feature_name = V2) 
feature_names <- features$feature_name
feature_names <- sub("^f", "frequency_", feature_names)
feature_names <- sub("^t", "time_", feature_names)
feature_names <- sub("Acc", "accelerometer_", feature_names)
feature_names <- sub("Gyro", "gyroscope_", feature_names)
feature_names <- sub("Mag", "magnitude_", feature_names)
feature_names <- sub("Gravity", "gravity_", feature_names)
feature_names <- sub("Jerk", "jerk_", feature_names)
feature_names <- sub("BodyBody", "body_", feature_names)
feature_names <- sub("Body", "body_", feature_names)
feature_names <- gsub("-", "", feature_names)
feature_names <- tolower(feature_names)
```

* Read in X_train.txt, comprised of readings from the training set 
* Rename the 561 columns in the 'X_train' dataframe using the 561 feature descriptions in 'feature_names' vector

```{r}
X_train <- read.table("./data/UCI HAR Dataset/train/X_train.txt", stringsAsFactors = F)
colnames(X_train) <- feature_names
```

* Read in X_test.txt, comprised of readings from the test set 
* Rename the 561 columns in the 'X_test' dataframe using the 561 feature descriptions in 'feature_names' vector

```{r}
X_test <- read.table("./data/UCI HAR Dataset/test/X_test.txt", stringsAsFactors = F) 
colnames(X_test) <- feature_names 
```

* Read in activity_labels.txt, which links the class labels with their activity name
* Rename the two columns in the 'activity_labels' dataframe as activity_no & activity_type

```{r}
activity_labels <- read.table("./data/UCI HAR Dataset/activity_labels.txt", stringsAsFactors = F)
activity_labels <- rename(activity_labels, activity_no = V1, activity_type = V2)
```

* Read in Y_train.txt, comprised of training labels
* Rename the lone column in the 'Y_train' dataframe as activity_no
* Left join the 'Y_train' dataframe to 'activity_labels' dataframe to add activity_type column using activity_no field as key 

```{r}
Y_train <- read.table("./data/UCI HAR Dataset/train/Y_train.txt", stringsAsFactors = F) 
Y_train <- rename(Y_train, activity_no = V1)
Y_train <- join(Y_train, activity_labels, by= "activity_no") 
```

* Read in Y_test.txt, comprised of test labels
* Rename the lone column in the 'Y_test' dataframe as activity_no
* Left join the 'Y_test' dataframe to 'activity_labels' dataframe to add activity_type column using activity_no field as key 

```{r}
Y_test <- read.table("./data/UCI HAR Dataset/test/Y_test.txt", stringsAsFactors = F) 
Y_test <- rename(Y_test, activity_no = V1)
Y_test <- join(Y_test, activity_labels, by= "activity_no")  
```

* Read in subject_train.txt, comprised of identifications of the 30 subjects who performed the activity for each window sample
* Rename the lone column in the 'subject_train' dataframe as subject_id

```{r}
subject_train <- read.table("./data/UCI HAR Dataset/train/subject_train.txt", stringsAsFactors = F) 
subject_train <- rename(subject_train, subject_id = V1) 
```

* Read in subject_test.txt, comprised of identifications of the 30 subjects who performed the activity for each window sample
* Rename the lone column in the 'subject_test' dataframe as subject_id

```{r}
subject_test <- read.table("./data/UCI HAR Dataset/test/subject_test.txt", stringsAsFactors = F) 
subject_test <- rename(subject_test, subject_id = V1) 
```

* Combine the three train dataframes 'X_train', 'Y_train', & 'subject_train' by column as 'cbind_train' dataframe
* Combine the three test dataframes 'X_test', 'Y_test', & 'subject_test' by column as 'cbind_test' dataframe
* Combine the two resulting dataframes 'cbind_train' & 'cbind_test' by row as 'rbind_dataset' dataframe

```{r}
cbind_train <- cbind(X_train, Y_train, subject_train) 
cbind_test <- cbind(X_test, Y_test, subject_test) 
rbind_dataset <- rbind(cbind_train, cbind_test) 
```

* Create vector, 'colgr' of column names that include including "mean()" and/or "std()" (but exclude "meanFreq()") from 'rbind_dataset' 

```{r}
colgr <- grep("mean\\()|std\\()", names(rbind_dataset), value = TRUE) 
```

* Subset the 'rbind_dataset' dataframe to include only columns identifying the subject and activity as well as those included in the 'colgr' vector

```{r}
subset_data <- rbind_dataset[ , c('subject_id', 'activity_type', colgr)] 
```

* Melt the resulting 'subset_data' dataframe using the columns identifying the subject and activity into 'subset_data_melt' datafram

```{r}
subset_data_melt <- melt(subset_data, id = c('subject_id', 'activity_type')) 
```

* Cast the resulting molten data frame into 'subset_data_mean' data frame to return average of each variable for each activity and subject

```{r}
subset_data_mean <- dcast(subset_data_melt, subject_id + activity_type ~ variable, mean)
```

* Export text file containing resulting data

```{r}
write.table(subset_data_mean, file = "TidyDataSet.txt", row.names = FALSE) 
```
}
```
